---
title: "Report on Controls"
author: "Emily Bean"
date: "March 25, 2020"
output: 
  rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 5, fig.width = 6, fig.align = "center", warning = FALSE, message = FALSE)

```


```{r}
# read in data
require(dplyr)
require(tibble)
require(stringr)
require(ggplot2)
dat <- read.table("https://github.com/EmilyB17/amr-brazil/blob/master/data/parsedRelativeAbundance.txt?raw=true", sep = "\t", header = TRUE)

# set theme
theme_set(theme_bw())

### filter only control samples
# pick out control samples
controls <- paste("G527_94_NoTemplate-DNAextraction2_S94",
                  "G527_96_NoTemplate-LibraryPrep_S96",
                  "G527_95_MockCommunity_S95",
                  "G527_93_NoTemplate-DNAextraction1_S93",
                  "G527_49_2019_DNA_S49", sep = "|")

# get dataframe of controls only
controlsDF <- dat %>% 
  filter(str_detect(name, controls)) %>% 
  # remove Alien sample
  filter(!str_detect(name, "G527_49_2019_DNA_S49"))

```

### Positive Control - Mock Community


```{r}

# get positive control
poscontrol <- controlsDF %>% filter(name == "G527_95_MockCommunity_S95") %>% 
  # remove zeros
  filter(relabun > 0)

```


There are `r length(poscontrol$gene)` genes detected in the mock community.

```{r}

# boxplot
ggplot(data = poscontrol, aes(x = class, y = relabun)) +
  geom_boxplot() +
  labs(x = "Resistance Class", y = "Relative Abundance", title = "Mock Community")

```


### Negative Control - DNA Extraction



```{r}

# get DNA extraction negative controls
dna <- paste("G527_93_NoTemplate-DNAextraction1_S93",
             "G527_94_NoTemplate-DNAextraction2_S94", sep = "|")
dnacontrol <- controlsDF %>% 
  mutate(name = as.character(name)) %>% 
  filter(str_detect(dna, name)) %>% 
  # get only present genes
  filter(relabun > 0) %>% 
  mutate(Control = factor(case_when(
    name %in% "G527_93_NoTemplate-DNAextraction1_S93" ~ 1,
    name %in% "G527_94_NoTemplate-DNAextraction2_S94" ~ 2
  )))

```

There are 2 DNA extraction controls. There are `r length(dnacontrol$gene[dnacontrol$Control == "1"])` genes detected in the 1st control and `r length(dnacontrol$gene[dnacontrol$Control == "2"])` detected in the 2nd control.

```{r}

# boxplot
ggplot(data = dnacontrol, aes(x = class, y = relabun)) +
  geom_boxplot() +
  labs(x = "Resistance Class", y = "Relative Abundance", title = "DNA Extraction Controls 1 & 2") +
  facet_wrap(~Control)

```

### Negative Control - Library Preparation



```{r}

# get library prep control
libcontrol <- controlsDF %>% 
  filter(name == "G527_96_NoTemplate-LibraryPrep_S96") %>% 
  # filter out zeros
  filter(relabun > 0)

```


There are only `r length(libcontrol$gene)` genes detected in this negative control.

```{r}

# boxplot
ggplot(data = libcontrol, aes(x = class, y = relabun)) +
  geom_boxplot() +
  labs(x = "Resistance Class", y = "Relative Abundance", title = "Library Prep Control")

```

