---
title: "DESeq"
author: "Emily Bean"
date: "9/14/2020"
output: 
  rmarkdown::github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
# set root directory b/c we're in Rmd
knitr::opts_knit$set(root.dir = '~/git/amr-heavymetal/')
```

## Differential abundance analysis at the genus level

This script reads in OTU tables generated in MG-RAST for farm A and farm B in fecal, rumen fluid, and DNS samples. Differential abundance analysis is performed at the genus level for differences between farms at each body site.

```{r read data}

## get data
require(tidyverse)
require(DESeq2)
require(phyloseq)
set.seed(123)

# set alpha
alpha = 0.05

# there are 3 text files, one for each anatomical site
# for DEseq format, need genus on rows and samples in columns

dns <- read.table("./data/OTU_genus_DNS_nominitable.txt", header = TRUE, sep = "\t") %>% 
  # pivot long by genus
  pivot_longer(cols = 2:last_col(), names_to = "G", values_to = "count") %>%
  # spread back out by sample name
  pivot_wider(names_from = genus, values_from = count, values_fill = 0) %>% 
  # get rid of NA
  drop_na() %>% 
  # make genus into rownames for DESeq
  column_to_rownames(var = "G")

# also needs to be in matrix format
dns <- as.matrix(dns)

fec <- read.table("./data/OTU_genus_feces.txt", header = TRUE, sep = "\t") %>% 
  # pivot long by genus
  pivot_longer(cols = 2:last_col(), names_to = "G", values_to = "count") %>%
  # spread back out by sample name
  pivot_wider(names_from = genus, values_from = count, values_fill = 0) %>% 
  # get rid of NA
  drop_na() %>% 
  # make genus into rownames for DESeq
  column_to_rownames(var = "G")

# also needs to be in matrix format
fec <- as.matrix(fec)

rum <- read.table("./data/OTU_genus_rumen.txt", header = TRUE, sep = "\t")  %>% 
# pivot long by genus
  pivot_longer(cols = 2:last_col(), names_to = "G", values_to = "count") %>%
  # spread back out by sample name
  pivot_wider(names_from = genus, values_from = count, values_fill = 0) %>% 
  # get rid of NA
  drop_na() %>% 
  # make genus into rownames for DESeq
  column_to_rownames(var = "G")

# also needs to be in matrix format
rum <- as.matrix(rum)

```


### DNS


DESeq requires a separate matrix with metadata (categorical information) and it needs to be structured properly for the janky software to work properly.

```{r metadata dns}

# get vector of sample names
samp <- colnames(as.data.frame(dns)) 

# turn this into a 1-column dataframe to add farm info
df <- as.data.frame(samp)

# add farm info that is stored in sample names
meta <- df %>% 
  # get string for farm
  mutate(farm = str_extract(samp, "P(\\d)")) %>% 
  # change to farm A and B
  mutate(farm = factor(case_when(
    farm %in% "P1" ~ "A",
    farm %in% "P2" ~ "B"
  ) )) %>% 
  # make into rownames
  column_to_rownames(var = "samp") 

# DESeq requires the reference level to be set 
meta$farm <- relevel(meta$farm, ref = "B")

# and it needs to be in matrix format
metam <- as.matrix(meta)

```

The DEseq protocol runs in the chunk below.

```{r deseq dns}

# because DESeq2 is janky,  the names have to be in THE EXACT ORDER
all(colnames(dns) == meta$sample)

# make DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = dns,
                              colData = metam,
                              design = ~ farm)
dds 

# perform DeSeq testing
de <- DESeq(dds, test = "Wald", fitType = "mean")  # fit type should be non-para

# get results
res = results(de, cooksCutoff = FALSE, contrast = c("farm", "A", "B"))
# get only significant results
sigtab = res[which(res$padj < alpha), ]

# sort by log2fold change
sigtab <- sigtab[order(sigtab$log2FoldChange), ]

# make dataframe to save as text table
tab <- as.data.frame(sigtab) %>% 
  rownames_to_column(var = "Genus")

# write results to file
#write.table(tab, file = "./data/DESeq-results-DNS.txt", sep = "\t", row.names = FALSE)

```

There are 297 significant genuses (geni?) out of 593 total in the DNS swabs between farm A and B.

### Rumen


DESeq requires a separate matrix with metadata (categorical information) and it needs to be structured properly for the janky software to work properly.

```{r metadata rumen}

# get vector of sample names
samp <- colnames(as.data.frame(rum)) 

# turn this into a 1-column dataframe to add farm info
df <- as.data.frame(samp)

# add farm info that is stored in sample names
meta <- df %>% 
  # get string for farm
  mutate(farm = str_extract(samp, "P(\\d)")) %>% 
  # change to farm A and B
  mutate(farm = factor(case_when(
    farm %in% "P1" ~ "A",
    farm %in% "P2" ~ "B"
  ) )) %>% 
  # make into rownames
  column_to_rownames(var = "samp") 

# DESeq requires the reference level to be set 
meta$farm <- relevel(meta$farm, ref = "B")

# and it needs to be in matrix format
metam <- as.matrix(meta)

```

The DEseq protocol runs in the chunk below.

```{r deseq rumen}

# because DESeq2 is janky,  the names have to be in THE EXACT ORDER
all(colnames(rum) == meta$sample)

# make DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = rum,
                              colData = metam,
                              design = ~ farm)
dds 

# perform DeSeq testing
de <- DESeq(dds, test = "Wald", fitType = "mean")  # fit type should be non-para

# get results
res = results(de, cooksCutoff = FALSE, contrast = c("farm", "A", "B"))
# get only significant results
sigtab = res[which(res$padj < alpha), ]

# sort by log2fold change
sigtab <- sigtab[order(sigtab$log2FoldChange), ]

# make dataframe to save as text table
tab <- as.data.frame(sigtab) %>% 
  rownames_to_column(var = "Genus")

# write results to file
#write.table(tab, file = "./data/DESeq-results-rumen.txt", sep = "\t", row.names = FALSE)

```
There are 50 significant genuses out of 600 in the rumen samples.


### Feces


DESeq requires a separate matrix with metadata (categorical information) and it needs to be structured properly for the janky software to work properly.

```{r metadata fecal}

# get vector of sample names
samp <- colnames(as.data.frame(fec)) 

# turn this into a 1-column dataframe to add farm info
df <- as.data.frame(samp)

# add farm info that is stored in sample names
meta <- df %>% 
  # get string for farm
  mutate(farm = str_extract(samp, "P(\\d)")) %>% 
  # change to farm A and B
  mutate(farm = factor(case_when(
    farm %in% "P1" ~ "A",
    farm %in% "P2" ~ "B"
  ) )) %>% 
  # make into rownames
  column_to_rownames(var = "samp") 

# DESeq requires the reference level to be set 
meta$farm <- relevel(meta$farm, ref = "B")

# and it needs to be in matrix format
metam <- as.matrix(meta)

```

The DEseq protocol runs in the chunk below.

```{r deseq fecal}

# because DESeq2 is janky,  the names have to be in THE EXACT ORDER
all(colnames(fec) == meta$sample)

# make DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = fec,
                              colData = metam,
                              design = ~ farm)
dds 

# perform DeSeq testing
de <- DESeq(dds, test = "Wald", fitType = "mean")  # fit type should be non-para

# get results
res = results(de, cooksCutoff = FALSE, contrast = c("farm", "A", "B"))
# get only significant results
sigtab = res[which(res$padj < alpha), ]

# sort by log2fold change
sigtab <- sigtab[order(sigtab$log2FoldChange), ]

# make dataframe to save as text table
tab <- as.data.frame(sigtab) %>% 
  rownames_to_column(var = "Genus")

# write results to file
#write.table(tab, file = "./data/DESeq-results-feces.txt", sep = "\t", row.names = FALSE)

```
There are 220 significant genuses out of 600 in the fecal samples.