---
title: "AMR Analysis"
author: "Emily Bean"
date: "March 12, 2020"
output: 
  rmarkdown::github_document:
      toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_knit$set(root.dir = "~/git/amr-heavymetal")

```

```{r}

require(dplyr)
require(tidyr)
require(stringr)
require(ggplot2)
require(ggpubr)

# read in data
counts <- read.table("./data/parsedRelativeAbundanceNoContams.txt", sep = "\t", header = TRUE)

# set ggplot2 theme
theme_set(theme_bw())

# get unique factors for loops
classes <- unique(counts$broadclass)
types <- unique(counts$type)
proteins <- unique(counts$protein)
genes <- unique(counts$gene)
patterns <- unique(counts$pattern)

# there were four samples that got mis-labeled; this needs to be fixed
### NOTE; this does not affect the FARM comparison b/c all were from farm2
wonkies <- paste("G527_66_P2A10_rumen_S66", "G527_67_P2A14_rumen_S67",
             "G527_79_P2A10_SNP_S79", "G527_80_P2A14_SNP_S80", sep = "|")

counts.fix <- counts %>% 
  # create new column of sites with the updated swaps
  mutate(site.fix = case_when(
    name %in% "G527_66_P2A10_rumen_S66" ~ "SNP",
    name %in% "G527_67_P2A14_rumen_S67" ~ "SNP",
    name %in% "G527_79_P2A10_SNP_S79" ~ "rumen",
    name %in% "G527_80_P2A14_SNP_S80" ~ "rumen",
    !str_detect(wonkies, name) ~ site
  ),
  farm = factor(farm))

```



## Analysis Structure

**Comparisons**  

Pairwise comparisons between Farm 1 and Farm 2.  
Comparisons between body sites (rumen, feces, nasal swab) at each farm.  

MEGARES database categories include: type of resistance, resistance class within type (i.e. type of drug within Drug resistance), protein class, and gene. Pairwise comparisons can be made for all of these but it seems to make the most biological sense to compare genes, protein classes, and resistance classes.  

***

Therefore, here is the proposed comparison structure:  
Genes: Farm 1 vs Farm 2  
Proteins: Farm 1 vs Farm 2  
Resistance class: Farm 1 vs Farm 2  
**Wilcox Rank Sum test**

Genes: SNP vs Rumen vs Feces (Farm 1 and Farm 2)  
Proteins: SNP vs Rumen vs Feces (Farm 1 and Farm 2)  
Resistance class: SNP vs Rumen vs Feces (Farm 1 and Farm 2)  
**Kruskal Wallis test**

***


### Data Exploration  

```{r}

# in the entire dataset, there are 147 unique proteins in 3 classes
exp <- counts %>% 
  group_by(broadclass, type, protein) %>% 
  summarize()

# show unique proteins
#knitr::kable(exp)

```

```{r}

# data is not normally distributed and is very zero-skewed
hist(counts$relabun, main = "Histogram of Relative Abundance",
     xlab = "Relative abundance", ylab = "Frequency")

```

### Pairwise comparisons between farms

Wilcox Rank-Sum test for non-parametric pairwise test of relative abundnace between farms

#### Genes

```{r}

# Wilcox rank-sum for pairwise differences between farms

# create empty df to fill in the loop
outsig <- data.frame()
# loop through all 666 genes
for(i in 1:length(patterns)) {
  
  # perform wilcox test
  w <- wilcox.test(relabun ~ farm, data = filter(counts, pattern == patterns[i]),
                   exact = FALSE)
  
  # get output p value
  outsig[i, "pattern"] <- patterns[i]
  outsig[i, "pval"] <- round(w$p.value, 3)
  
}

# adjust p values for false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))

# get only significant variables and add median + IQR
sigs <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = "pattern") %>% 
  # add presence/absence
    mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(farm, pattern, broadclass, type, protein, gene) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
  # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  
  ungroup() %>% 
  mutate(farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  pivot_wider(names_from = farm, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  # add p value
  left_join(adj, by = "pattern") %>% 
  select(-c(pattern, pval, sig))
#write.table(sum, "/Users/epb5360/git/amr-brazil/data/results-tables/wilcox-genes.txt", sep = "\t", row.names = FALSE)

```

There are `r length(unique(sigs$pattern))` genes that are significantly different between farms.


Visualize with presence/absence:

```{r}

# plot by presence instead of relative abundance
ggbarplot(sigs, x = "gene", y = "prestotal",
          fill = "farm",
          position = position_dodge(0.9)) +
  labs(x = "Gene", y = "Sample presence")

```



```{r}

## plot all data (not just  median)

# get significant data
allsig <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = "pattern") %>% 
  mutate(farm = factor(farm))

# plot
ggplot(data = allsig, aes(x = farm, y = relabun,fill = farm)) +
  geom_violin() +
  geom_point(position = position_dodge2(width = 0.5), size = 0.3) +
  facet_grid(pattern~broadclass, scales = "free") +
  labs(x = "Farm", y = "Relative Abundance", title = "Relative abundance of significant genes")

```




#### Proteins

```{r}
outsig <- data.frame()
for(i in 1:length(proteins)) {
  
  w <- wilcox.test(relabun ~ farm, data = filter(counts, protein == proteins[i]),
                   exact = FALSE)
  
  outsig[i, "protein"] <- proteins[i]
  outsig[i, "pval"] <- round(w$p.value, 3)
  
}

# adjust p values for false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))

# get only significant variables and add median + IQR
sigs <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = "protein") %>% 
  # add presence/absence
    mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(farm, broadclass, type, protein) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
  # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  
  ungroup() %>% 
  mutate(farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  pivot_wider(names_from = farm, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  # add p value
  left_join(adj, by = "protein") %>% 
  select(-c(pval, sig))
#write.table(sum, "/Users/epb5360/git/amr-brazil/data/results-tables/wilcox-proteins.txt", sep = "\t", row.names = FALSE)

```

There are `r length(unique(sigs$protein))` significantly different resistance proteins; the plot shows that most are only present in one of the two farms.


Visualize with presence/absence:

```{r}

# plot by presence/absence
ggplot(data = sigs, aes(x = protein, y = prestotal, fill = farm)) +
  geom_col(position = position_dodge(0.9)) +
  facet_wrap(~broadclass, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



There are 16 proteins with significant differences, so that's difficult to visualize on one plot. 

```{r}


## plot all data (not just  median)

# get significant data
allsig <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = "protein") %>% 
  mutate(farm = factor(farm))

# plot
ggplot(data = allsig, aes(x = farm, y = relabun, fill = farm)) +
  geom_violin() +
  geom_point(position = position_dodge2(width = 0.5), size = 0.3) +
  facet_wrap(~protein) +
  labs(x = "Farm", y = "Relative Abundance", title = "Relative abundance of significant proteins")

```

Visualize with density plots; this shows that most of the proteins have marginally higher relative abundance in farm 1.

```{r}

# get significant data
sigrel <- counts %>% 
  semi_join(filter(adj, padj < 0.05), by = "protein") %>% 
  mutate(farm = factor(farm))

# plot with ggpubr
ggdensity(sigrel, x = "relabun", y = "..count..", add = "median", 
          color = "farm", fill = "farm", rug = TRUE) +
  facet_wrap(~protein, scales = "free")

```

#### Resistance Type

```{r}


outsig <- data.frame()
for(i in 1:length(types)) {
  
  w <- wilcox.test(relabun ~ farm, data = filter(counts, type == types[i]),
                   exact = FALSE)
  
  outsig[i, "type"] <- types[i]
  outsig[i, "pval"] <- round(w$p.value, 3)
  
}

# adjust p values for false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))

# get only significant variables and add median + IQR
sigs <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = "type") %>% 
  # add presence/absence
    mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(farm, broadclass, type) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
   # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  
  ungroup() %>% 
  mutate(farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  pivot_wider(names_from = farm, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  # add p value
  left_join(adj, by = "type") %>% 
  select(-c(pval, sig))
#write.table(sum, "/Users/epb5360/git/amr-brazil/data/results-tables/wilcox-types.txt", sep = "\t", row.names = FALSE)

```

There are `r length(unique(sigs$type))` significantly different resistance types.

```{r}

# Visualize with presence/absence
ggplot(data = sigs, aes(x = type, y = prestotal, fill = farm)) +
  geom_col(position = position_dodge(0.9)) +
  facet_wrap(~broadclass, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




```{r}

## plot all data (not just  median)

# get significant data
allsig <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = "type") %>% 
  mutate(farm = factor(farm))

# plot
ggplot(data = allsig, aes(x = farm, y = relabun, fill = farm)) +
  geom_violin() +
  geom_point(position = position_dodge2(width = 0.5), size = 0.3) +
  facet_wrap(~type) +
  labs(x = "Farm", y = "Relative Abundance", title = "Relative abundance of significant resistance type")


```


#### Resistance Class

```{r}

outsig <- data.frame()
for(i in 1:length(classes)) {
  
  w <- wilcox.test(relabun ~ farm, data = filter(counts, broadclass == classes[i]),
                   exact = FALSE)
  
  outsig[i, "class"] <- classes[i]
  outsig[i, "pval"] <- round(w$p.value, 3)
  
}

# adjust p values for false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))

# get only significant variables and add median + IQR
sigs <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = c("class" = "broadclass")) %>% 
  # add presence/absence
    mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(farm, class) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
     # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  ungroup() %>% 
  mutate(farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  pivot_wider(names_from = farm, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  # add p value
  left_join(adj, by = "class") %>% 
  select(-c(pval, sig))
#write.table(sum, "/Users/epb5360/git/amr-brazil/data/results-tables/wilcox-class.txt", sep = "\t", row.names = FALSE)

```

There are `r length(unique(sigs$class))` significantly different resistance types.

```{r}

# Visualize with presence/absence
ggplot(data = sigs, aes(x = class, y = prestotal, fill = farm)) +
  geom_col(position = position_dodge(0.9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



```{r}

## plot all data (not just  median)

# get significant data
allsig <- adj %>% 
  filter(sig == "TRUE") %>% 
  left_join(counts, by = c("class" = "broadclass")) %>% 
  mutate(farm = factor(farm))

# plot
ggplot(data = allsig, aes(x = farm, y = relabun, fill = farm)) +
  geom_violin() +
  geom_point(position = position_dodge2(width = 0.5), size = 0.3) +
  facet_wrap(~type) +
  labs(x = "Farm", y = "Relative Abundance", title = "Relative abundance of significant resistance class")

```

### Comparisons between body sites

Kruskal-Wallis test for non-parametric comparisons between the 3 body sites, stratified by farm.

#### Genes

```{r}


## Stratify for each farm
farms <- levels(counts.fix$farm)
# create empty dataframe to fill within the loop
outsig <- data.frame()

for(j in 1:length(farms)) {
  
  # iterate through all 666 genes
  for(i in 1:length(patterns)) {
    
    # perform Kruskal-Wallis test
    k <- kruskal.test(relabun ~ site.fix, data = filter(counts.fix, pattern == patterns[i] &
                                                      farm == farms[j]))
    
   
    # some farms do not have the gene present; mark as 'absent'
    if(is.na(k$p.value)) {
      
      pvaldat <- NA
      
    } else {
      
      pvaldat <- round(k$p.value, 3)
      
    } 
        
    # collect variables to output DF
    outsig <- rbind(outsig,
                    data.frame(farm = farms[j],
                               pattern = patterns[i],
                               pval = pvaldat))
  }
}

# adjust p vals for multiple comparisons with false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))
 
# get genes that had a significant difference
sigs <- adj %>% filter(sig == "TRUE") %>%  
  left_join(counts.fix, by = c("pattern", "farm")) %>% 
  # add presence/absence
  mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(site.fix, farm, broadclass, type, protein, gene, pattern) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
      # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  ungroup() %>% 
  mutate(site.fix = factor(site.fix),
         farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  left_join(adj, by = c("pattern", "farm")) %>% 
  select(-c(pval, sig)) %>% 
  pivot_wider(names_from = site.fix, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  select(-pattern) %>% 
  arrange(farm)
  
#write.table(sum, "./data/kruskal-genes.txt", sep = "\t", row.names = FALSE)

```

There are `r length(unique(sigs$pattern))` significant genes between body site.fixs. The p values and summary statistics are not printed since there are so many.

Therefore, it may be more helpful to analyze this data in terms of the fewer genes that had higher prevalence in rumen or SNP, than fecal.

```{r}

## confirm if there are any standout genes that are higher in either rumen or SNP compared to fecal
s <- sum %>% 
  select(-c(starts_with("median"), starts_with("IQR"), padj, starts_with("persamp"))) %>% 
  # ssubtract fecal from SNP and rumen samples
  # are fecal samples always higher in AMR prevalence?
  mutate(diff.rumfec = prestotal_rumen - prestotal_feces,
         diff.SNPfec = prestotal_SNP - prestotal_feces) 

# get samples that are higher in DNS than feces
hiDNS <- s %>% 
  filter(diff.SNPfec > 0) # only 7
# how many on each farm?
length(which(hiDNS$farm == 1)) # 3
length(which(hiDNS$farm == 2)) # 4

# get samples that are higher in rumen than feces
hiRUM <- s %>% 
  filter(diff.rumfec > 0) # only 88
# how many on each farm?
length(which(hiRUM$farm == 1)) # 72
length(which(hiRUM$farm == 2)) # 16

## total for each farm
length(which(hiDNS$farm == 1))  + length(which(hiRUM$farm == 1)) # 75
length(which(hiDNS$farm == 2)) + length(which(hiRUM$farm == 2)) # 20

```




This plot visualizes the differences; most are higher in feces, then rumen, then SNP. However, one gene is higher in SNP.

```{r}

# visualize the major trends
ggplot(data = sigs, aes(x = site.fix, y = median, group = pattern)) +
  geom_point() +
  geom_jitter() +
  geom_line() +
  theme_bw() +
  labs(x = "Body site.fix", y = "Median Relative Abundance") +
  ggtitle("Significantly different genes by site.fix & farm",
          subtitle = paste0("n = (", length(which(adj$sig == TRUE)), ")")) +
  facet_wrap(~ farm)

```


#### Proteins

```{r}

## Stratify for each farm
counts.fix <- counts.fix %>% mutate(farm = factor(farm))
farms <- levels(counts.fix$farm)
# create empty dataframe to fill within the loop
outsig <- data.frame()

for(j in 1:length(farms)) {
  
  # iterate through all 666 genes
  for(i in 1:length(proteins)) {
    
    # perform Kruskal-Wallis test
    k <- kruskal.test(relabun ~ site.fix, data = filter(counts.fix, protein == proteins[i] &
                                                      farm == farms[j]))
    
   
    # some farms do not have the gene present; mark as 'absent'
    if(is.na(k$p.value)) {
      
      pvaldat <- NA
      
    } else {
      
      pvaldat <- round(k$p.value, 3)
      
    } 
        
    # collect variables to output DF
    outsig <- rbind(outsig,
                    data.frame(farm = farms[j],
                               protein = proteins[i],
                               pval = pvaldat))
  }
}


# adjust p vals for multiple comparisons with false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))
 
# get genes that had a significant difference
sigs <- adj %>% filter(sig == "TRUE") %>%  
  left_join(counts.fix, by = c("protein", "farm")) %>% 
  # add presence/absence
  mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(site.fix, farm, broadclass, type, protein) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
      # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  ungroup() %>% 
  mutate(site.fix = factor(site.fix),
         farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  left_join(adj, by = c("protein", "farm")) %>% 
  select(-c(pval, sig)) %>% 
  pivot_wider(names_from = site.fix, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  arrange(farm)
 
#write.table(sum, "./data/kruskal-proteins.txt", sep = "\t", row.names = FALSE)


```


There are `r length(unique(sigs$protein))` significant proteins in the dataset.


```{r}

# visualize the major trends
ggplot(data = sigs, aes(x = site.fix, y = median, group = protein)) +
  geom_point() +
  geom_jitter() +
  geom_line() +
  theme_bw() +
  labs(x = "Body site.fix", y = "Median Relative Abundance") +
  ggtitle("Significantly different proteins by site.fix & farm",
          subtitle = paste0("n = (", length(which(adj$sig == TRUE)), ")")) +
  facet_wrap(~ farm)

```


#### Resistance Type

```{r}


## Stratify for each farm
counts.fix <- counts.fix %>% mutate(farm = factor(farm))
farms <- levels(counts.fix$farm)
# create empty dataframe to fill within the loop
outsig <- data.frame()

for(j in 1:length(farms)) {
  
  # iterate through all 666 genes
  for(i in 1:length(types)) {
    
    # perform Kruskal-Wallis test
    k <- kruskal.test(relabun ~ site.fix, data = filter(counts.fix, type == types[i] &
                                                      farm == farms[j]))
    
   
    # some farms do not have the gene present; mark as 'absent'
    if(is.na(k$p.value)) {
      
      pvaldat <- NA
      
    } else {
      
      pvaldat <- round(k$p.value, 3)
      
    } 
        
    # collect variables to output DF
    outsig <- rbind(outsig,
                    data.frame(farm = farms[j],
                               type = types[i],
                               pval = pvaldat))
  }
}

# adjust p vals for multiple comparisons with false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))
 
# get genes that had a significant difference
sigs <- adj %>% filter(sig == "TRUE") %>%  
  left_join(counts.fix, by = c("type", "farm")) %>% 
  # add presence/absence
  mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(site.fix, farm, broadclass, type) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
     # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  ungroup() %>% 
  mutate(site.fix = factor(site.fix),
         farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  left_join(adj, by = c("type", "farm")) %>% 
  select(-c(pval, sig)) %>% 
  pivot_wider(names_from = site.fix, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  arrange(farm)
 
#write.table(sum, "./data/kruskal-types.txt", sep = "\t", row.names = FALSE)

```


There are `r length(unique(sigs$type))` significant proteins in the dataset.




```{r}

# visualize the major trends
ggplot(data = sigs, aes(x = site.fix, y = median, group = type)) +
  geom_point() +
  geom_jitter() +
  geom_line() +
  theme_bw() +
  labs(x = "Body site.fix", y = "Median Relative Abundance") +
  ggtitle("Significantly different types by site.fix & farm",
          subtitle = paste0("n = (", length(which(adj$sig == TRUE)), ")")) +
  facet_wrap(~ farm)

```


#### Class

```{r}


## Stratify for each farm
counts.fix <- counts.fix %>% mutate(farm = factor(farm))
farms <- levels(counts.fix$farm)
# create empty dataframe to fill within the loop
outsig <- data.frame()

for(j in 1:length(farms)) {
  
  # iterate through all 666 genes
  for(i in 1:length(classes)) {
    
    # perform Kruskal-Wallis test
    k <- kruskal.test(relabun ~ site.fix, data = filter(counts.fix, broadclass == classes[i] &
                                                      farm == farms[j]))
    
   
    # some farms do not have the gene present; mark as 'absent'
    if(is.na(k$p.value)) {
      
      pvaldat <- NA
      
    } else {
      
      pvaldat <- round(k$p.value, 3)
      
    } 
        
    # collect variables to output DF
    outsig <- rbind(outsig,
                    data.frame(farm = farms[j],
                               class = classes[i],
                               pval = pvaldat))
  }
}

# adjust p vals for multiple comparisons with false discovery rate
adj <- outsig %>% 
  mutate(
    # adjust p values with FDR
    padj = p.adjust(pval, method = "fdr", n = length(outsig$pval)),
    # Boolean for significance
    sig = case_when(
      padj < 0.05 ~ TRUE,
      padj >= 0.05 ~ FALSE))
 
# get genes that had a significant difference
sigs <- adj %>% filter(sig == "TRUE") %>%  
  left_join(counts.fix, by = c("class" = "broadclass", "farm")) %>% 
  # add presence/absence
  mutate(presence = case_when(
    relabun == 0 ~ 0,
    relabun != 0 ~ 1
  )) %>% 
  group_by(site.fix, farm, class) %>% 
  summarize(median = median(relabun),
            IQR = IQR(relabun),
            prestotal = sum(presence)) %>% 
      # calculate presence/absence per sample since groups had unequal sample size 
  mutate(persamp.pres = case_when(
    farm == 1 ~ prestotal / 48,
    farm == 2 ~ prestotal /43
  )) %>% 
  ungroup() %>% 
  mutate(site.fix = factor(site.fix),
         farm = factor(farm))

## write summary statistics to table
sum <- sigs %>% 
  left_join(adj, by = c("class", "farm")) %>% 
  select(-c(pval, sig)) %>% 
  pivot_wider(names_from = site.fix, values_from = c(median, IQR, prestotal, persamp.pres)) %>% 
  arrange(farm)
#write.table(sum, "./data/kruskal-class.txt", sep = "\t", row.names = FALSE)


```


There are `r length(unique(sigs$class))` significant classes in the dataset.



```{r}

# visualize the major trends
ggplot(data = sigs, aes(x = site.fix, y = median, group = class)) +
  geom_point() +
  geom_jitter() +
  geom_line() +
  theme_bw() +
  labs(x = "Body site.fix", y = "Median Relative Abundance") +
  ggtitle("Significantly different classes by site.fix & farm",
          subtitle = paste0("n = (", length(which(adj$sig == TRUE)), ")")) +
  facet_wrap(~ farm)

```